---
title: "Hypertension and microbiome"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: html_document
---
	
```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, xecho = TRUE, message = FALSE, results='asis', cache=FALSE, warning=FALSE)
knitr::opts_chunk$set(cache.path = 'cache/', output.dir="cache/",
                      file.path = 'cache/', fig.path = 'cache/')

dir.create("rds/", showWarnings = FALSE)
```

# Command line arguments

Calculations ran at

```{r Command line arguments}
now <- format(Sys.time(), '%Y%m%d-%H%M%S')

if (exists("args")) {
    if ("time" %in% names(args)) now <- args$time
    
    if("clean" %in% names(args)) {
        unlink("rds", recursive=TRUE)
        unlink("cache", recursive=TRUE)
    }
}
```

# Importing libraries:

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(microbiome)
library(finriskmetagcommon)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(ggpmisc)
library(Maaslin)
library(parallel)
library(officer)
library(flextable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
```

Session info

```{r Session info}
pander(sessionInfo(), compact = TRUE)
```

```{r Initialize docx, include = FALSE, cache = FALSE}
doc <- read_docx(path = "style/articlestyle.docx") %>%
  body_remove()
dir.create("rds", showWarnings = FALSE)
dir.create("cache", showWarnings = FALSE)
```

# Sources

### Officer functions

```{r Officer importer, code=readLines("articletwo-officer.R")}
source("articletwo-officer.R")
```

### RR biome functions

```{r RR biome functions, code=readLines("articletwo-rrbiome.R")}
source("articletwo-rrbiome.R")
```

### EdgeR plot functions

```{r EdgeR functions, code=readLines("aaro/edgeR_plot_functions.R")}
source("aaro/edgeR_plot_functions.R")
```

### Plot functions

```{r Plot functions, code=readLines("articletwo-ggplot.R")}
source("articletwo-ggplot.R")
```

# Loading data
	
Loading descriptions for clinical data

```{r variables, warning = FALSE}
names.dset <- getdescriptions()
```

Loading phyloseq object

```{r data}
pseq.genus <- readRDS("data/phfinrisk_genus_all_drop50k_2018-11-16.RDs")
pseq.species <- readRDS("data/phfinrisk_species_all_drop50k_2018-12-21.RDs")
```

Updating phyloseq objects

```{r updating phyloseq objects}
phyloseq::sample_data(pseq.genus) <- phyloseq::sample_data(filter.phenotype.data(pseq.genus))
phyloseq::sample_data(pseq.species) <- phyloseq::sample_data(filter.phenotype.data(pseq.species))
```

Sample data has dimensions (`r dim(sample_data(pseq.species))`).

## Models

```{r matrix calculation}
if (!file.exists("rds/bray.dist.m.species.rds")) {
    bray.dist.m.species <- calculate.beta.matrix(pseq.species)
    saveRDS(bray.dist.m.species, file = "rds/bray.dist.m.species.rds")
} else {
    bray.dist.m.species  <- readRDS("rds/bray.dist.m.species.rds")
}
```

 ```{r adonis calculation}
 if (!file.exists("rds/adonis.species.rds")) {
     adonis.species = list()
     adonis.species$min <- calculateadonis(dset = meta(pseq.species),
                                           matrix = bray.dist.m.species,
                                           covariates = c("BL_AGE", "SEX"),
                                           npermutations = 1000)
     adonis.species$max <- calculateadonis(dset = meta(pseq.species),
                                           matrix = bray.dist.m.species,
                                           covariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB",
                                                          "BL_USE_RX_C03", "BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09"),
                                           npermutations = 1000)
     saveRDS(adonis.species, file = "rds/adonis.species.rds")
    } else {
     adonis.species  <- readRDS("rds/adonis.species.rds")
}
```

```{r pca calculate}
if (!file.exists("rds/pca.prcomp.rds")) {
    pca.abundances  <- as.matrix(t(microbiome::abundances(microbiome::transform(pseq.species, "clr"))))
    pca.prcomp <- stats::prcomp(pca.abundances)
    saveRDS(pca.prcomp, file = "rds/pca.prcomp.rds")
} else {
    pca.prcomp <- readRDS("rds/pca.prcomp.rds")
}
```

1 %, 0.1 %

```{r maaslin runs, results = 'hide'} 
maaslin <- list()

#core.taxa.standard <- core_members(pseq.rel, detection = 0, prevalence = 50/100)

maaslin$genus <- maaslinwrapper(pseq = pseq.genus,
                looped = c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION"),
                forced = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB", "BL_USE_RX_C03",
                           "BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09"),
                taxa = "Halapricum.Archaea-")

lactotaxa <- paste0(c("Lactobacillus_delbrueckii.Bacteria", "Lactobacillus_ruminis.Bacteria", "Lactobacillus_acidophilus.Bacteria",
                    "Lactobacillus_fermentum.Bacteria", "Lactobacillus_paracasei.Bacteria", "Lactobacillus_salivarius.Bacteria"), collapse = ",")

maaslin$lacto.min <- maaslinwrapper(pseq = (subset_taxa(pseq.species, Genus=="Lactobacillus")),
                looped = c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION"),
                forced = c("BL_AGE", "SEX"),
                taxa = lactotaxa)

maaslin$lacto.max  <- maaslinwrapper(pseq = (subset_taxa(pseq.species, Genus=="Lactobacillus")),
                looped = c("MAP", "SYSTM", "DIASM", "PULSEPRESSURE", "HYPERTENSION"),
                forced =c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "DIAB", "BL_USE_RX_C03",
                          "BL_USE_RX_C07", "BL_USE_RX_C08", "BL_USE_RX_C09"),
                taxa = lactotaxa)
```

## Characteristics of the study sample. (Table 1.)

```{r Characteristics}
tbl1 <- tableone(meta(pseq.species))
```

```{r Write characteristics}
tbl1head <- "Characteristics of the study sample."
tbl1foot <- "Continuous variables are presented as mean (standard deviation). Exercise at free-timeis reported as not much (1), walk etc. (2), regular three hours or more or competitive sports (3). Categorical variables reported as absolute and relative frequencies. BP indicates blood pressure, BMI body mass index, RAS renin-angiotensin system."
writetable(doc, tbl1, number = 1, tbl1head, tbl1foot)
```

## Alpha & Beta plot (Figure 1.)

```{r alphabeta definitions}
dset.alpha <- meta.merge.alphadiversity(pseq.species)
alphadiversity <- list(
    min = calculateglm(dset.alpha, modelstring = "%s ~ BL_AGE + SEX  + diversities_shannon"),
    max = calculateglm(dset.alpha, modelstring = "%s ~ BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09  + diversities_shannon"))
diversities <- mergediversities(alphadiversity, adonis.species)
```

```{r save grob}
alpha.diversity.min <- plot.alpha(diversities$min, colorlimit = 1.3)
alpha.diversity.max <- plot.alpha(diversities$max, colorlimit = 1.3)
beta.diversity.min <- plot.beta(diversities$min)
beta.diversity.max <- plot.beta(diversities$max)


gs4 <- list(alpha.diversity.min$yaxis,
            alpha.diversity.min$plot,
            beta.diversity.min$plot,  
            alpha.diversity.min$legend,
            alpha.diversity.min$xaxis,
            beta.diversity.min$xaxis,
            beta.diversity.min$xlab,
            text_grob("A.", size = 18),
            alpha.diversity.max$yaxis,
            alpha.diversity.max$plot,
            beta.diversity.max$plot,  
            alpha.diversity.max$xaxis,
            beta.diversity.max$xaxis,
            beta.diversity.max$xlab,
            text_grob("B.", size = 18))

g4 <- arrangeGrob(grobs = gs4,
                  layout_matrix = rbind(
                      c(8,    NA,    NA, NA,   NA, NA, NA),
                      c(NA, 1,    2,    NA, 3, NA, NA),
                      c(NA, 1,    2,    NA, 3, NA, 4),
                      c(NA, NA, 5,    NA,6, NA, 4),
                      c(NA, NA, NA, NA,7, NA, 4),
                      c(15,    NA,    NA, NA,   NA, NA, 4),
                      c(NA, 9,    10,  NA,  11, NA, 4),
                      c(NA, 9,    10,  NA,  11, NA, NA),
                      c(NA, NA, 12,   NA, 13, NA, NA),
                      c(NA, NA, NA, NA,14, NA, NA)),
                  widths=c(0.7, 4, 0.7, 0.1, 5, 0.7, 4),
                  heights=rep(c(1, 1, 6, 1, 2), 2))
ggsave(file = "cache/alpha-beta.png", plot=g4, height=6, width=9)
```

```{r Write figure 1}
fig1head <- "alpha beta figure"
fig1foot <- ""
writeimage(doc, 1, "cache/alpha-beta.png", fig1head, fig1foot)
```

## Principal component analysis (Figure 2.)

```{r pca plot}
pca.eigs <- pca.prcomp$sdev^2
pca.percentage <- 100 * pca.eigs /sum(pca.eigs)
pca.labels <- sprintf("%s (%.2f %%)", colnames(pca.prcomp$x), pca.percentage)

pcaplot <- ggplot(cbind(meta(pseq.species), pca.prcomp$x[,1:2]), aes(x=PC1, y=PC2)) +
    geom_point(aes(color = HYPERTENSION), size = 0.5) +
    scale_color_manual(values=c("blue", "red"),
                       name= "Hypertension",
                       breaks=c("1", "0"),
                       labels=c("Yes", "No")) +
    xlab(pca.labels[1]) +
    ylab(pca.labels[2]) +
    coord_fixed() +
    scale_x_continuous(breaks=seq(-100, 100, 20)) +
    scale_y_continuous(breaks=seq(-30, 30, 10)) +
    theme_classic() +
    theme(text = element_text(size=20),
          legend.justification=c(1,0),
          legend.position=c(1.03, 0.75),
          legend.title=element_text(size = 20),
          legend.direction="vertical",
          legend.text=element_text(size = floor(0.75*20)),
          legend.box.margin = margin(1, 1, 1, 1, "mm"),
          legend.box.background = element_rect(colour = "black"))

ggsave(file = "cache/pca.png", plot = pcaplot, height = 5, width = 9, dpi = 300)
```

```{r Write figure 2}
fig2head <- "Tenth of bacterial variation can be explained by two first principal axes"
fig2foot <- sprintf("Principal component analysis for bacterial abundances at species level. Centered log ratio transform is performed for the compositional data. First two axes explain %.1f%% of the variation. Research participants with hypertension are denoted by red dot.", sum(pca.percentage[1:2]))
writeimage(doc, 2, "cache/pca.png", fig2head, fig2foot)
```

## Direct associations between genus level and blood pressure variables (figure 3)

```{r maaslin}
maas.results.df.signf <- maaslin$genus %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
           beta = Coefficient) %>%
   dplyr::filter(Qval < 0.05) 

maaslinplot <- plot_maaslin_heatmap_final(maas.results.df.signf %>% dplyr::filter(Variable != "HYPERTENSION"),
                      beta.maxabs=max(abs(maas.results.df.signf[,'beta'])),
                      order.by.category=FALSE)

ggsave(file = "cache/maaslin.png", plot = maaslinplot, height = 3, width = 7, dpi = 300)
```

```{r Write figure 3}
fig3head <- ""
fig3foot <- ""
writeimage(doc, 3, "cache/maaslin.png", fig3head, fig3foot)
```


## Figure 4

```{r PC plot}
pca.prcomp.df <- pca.prcomp$x[,1:5] %>% data.frame
#pca.prcomp$sdev[1:20]^2/sum(pca.prcomp$sdev^2)
#cumsum(pca.prcomp$sdev[1:20]^2/sum(pca.prcomp$sdev^2))

X <- base::merge(meta(pseq.species), pca.prcomp.df, by=0, all=TRUE) %>%
    dplyr::rename(Sample_ID = Row.names) 

X.res <- lapply(vectortolist(paste0("PC", seq(5))), function(axis) {
    modelstring <- paste0("%s ~  BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09 + ", axis)
    calculateglm(X, modelstring = modelstring, filterstr = "PC")
})

Z.df <-  do.call(rbind.data.frame, X.res) %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="response", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(p.value, method="BH"),
                  beta = estimate,
                  Feature = term)

p2 <- plot_maaslin_heatmap_final(Z.df , fname="Feature", pname="Name",
                                  xlab.text = "PC", signif="Qval", cluster=FALSE,
                                  v.title.name = "Effect size (PC)",
                                  ystr="reorder(Name, Qval)", sign.name="Significance.PC",
                                  no.sign.legend=TRUE)

ggsave(file = "cache/pca.p2.png", plot = p2, height = 8, width = 8, dpi = 300) 
```


```{r Write figure 4}
fig4head <- "P2"
fig4foot <- ""
writeimage(doc, 4, "cache/pca.p2.png", fig4head, fig4foot)
```


## Figure 5

```{r PCA heatmap}
pca.axis <- pca.prcomp$x %>% data.frame
pca.dset <- base::merge(meta(pseq.species), pca.axis[,1:5], by=0, all=TRUE) %>%
  dplyr::rename(Sample_ID = Row.names)

pca.dset %>% head

pca.ret <- lapply(paste0("PC", seq(1,5)), function (pc) {
    calculateglm(pca.dset,
                 modelstring = paste("%s ~ BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09  +", pc),
                 filterstr = "PC") %>%
        select(response, estimate, p.value) %>%
        mutate(model_name = pc)
})

pca.ret  %>%
    data.table::rbindlist() %>%
    as.data.frame %>%
    mutate(p.value = p.adjust(p.value, method="BH")) %>%
    arrange(p.value) %>%
    kable
```

## Loadings for principal component axis 3 (Figure 5)

```{r Main contributors of PC3}
pca.rotation.abs <- abs(pca.prcomp$rotation)
pca.loading <- sweep(pca.rotation.abs, 2, colSums(pca.rotation.abs), "/") %>%
	    as.data.frame %>%
    tibble::rownames_to_column(var = "term") %>%
    mutate(name = rename.genus(term))
	


pca.loading.pc3.top20 <- pca.loading %>%
    select(name, PC3) %>%
    arrange(-PC3) %>%
    head(20)

pca.pc3.gg <- ggplot(data=pca.loading.pc3.top20, aes(x=reorder(name, -PC3), y=100*PC3)) +
    geom_bar(stat="identity", width=0.5, position = position_dodge(preserve = "single")) +
    theme_classic(20) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.3, hjust = 1)) +
    xlab("") +
    ylab("Contribution (%)") +
    scale_y_continuous(limits=c(0, 2.7), oob = rescale_none, expand = c(0,0)) +
    theme(axis.title.x=element_blank(),
          text = element_text(size=20),
          legend.justification=c(1,0),
          legend.position=c(1.03, 0.75),
          legend.title=element_text(size = 20),
          legend.direction="vertical",
          legend.text=element_text(size = floor(0.75*20)),
          legend.box.margin = margin(1, 1, 1, 1, "mm"),
          legend.box.background = element_rect(colour = "black"),
          aspect.ratio = 0.7)
   ggsave(file = "cache/pca.pc3.png", plot = pca.pc3.gg, height = 8, width = 8, dpi = 300) 
```

```{r Write figure 5}
fig5head <- "Bacteria causing dysentery and diarrhea contribute to third principal component"
fig5foot <- sprintf("Top 20 bacterial contributors to third principal coordinate at species level. Centered log ratio transform is performed for the compositional data. Third principal component explain %.1f%% of the total variation. Plasmid DNA is denoted using star.", pca.percentage[3])
writeimage(doc, 5, "cache/pca.pc3.png", fig5head, fig5foot)
```

### Lactobacillus

Lacto min

```{r Reading maaslin lacto min output}
maas.lacto.results.df.signf <- maaslin$lacto.min %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  beta = Coefficient) #%>%

maas.lacto.results.df.signf %>%
    select(Variable, Feature, P.value, Q.value) %>%
    arrange(Q.value) %>%
    kable
```

Lacto max

```{r Reading maaslin lacto max output}
maas.lacto.results.df.signf <- maaslin$lacto.max %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  beta = Coefficient) #%>%

maas.lacto.results.df.signf %>%
    select(Variable, Feature, P.value, Q.value) %>%
    arrange(Q.value) %>%
    kable
```


```{r Write docx to file}
print(doc, target = paste0("report/articletwo-", now, ".docx"))
```

```{r save session}
save.image(file = paste0("cache/session-", now, ".Rdata"))
```
